---
title: "Ch. 12 Exercises"
author: "Cassandra"
date: "August 19, 2019"
output: 
  html_document: 
    keep_md: yes
---

```{r include = FALSE}  
knitr::opts_chunk$set(cache = TRUE, autodep = TRUE)
```

```{r message = FALSE}
library(tidyverse)
```

Joins tutorial (http://malooflab.phytonetworks.org/apps/joins-tutorial/)

## 12.2 Tidy data

```{r}
table1
table2
table3
table4a
table4b
```

The sample tibbles are included in the `tidyverse`, but this is one way to create those manually in R.
```
table1 <- tibble(
  country = rep(c("Afghanistan", "Brazil", "China"), each = 2),
  year = rep(c(1999, 2000), times = 3),
  cases = c(745, 2666, 37737, 80488, 212258, 213766),
  population = c(19987071, 20595360, 172006362, 174504898, 1272915272, 1280428583)
)

table2 <- tibble(
  country = rep(c("Afghanistan", "Brazil", "China"), each = 4),
  year = rep(c(1999, 2000), each = 2, times = 3),
  type = rep(c("cases", "population"), times = 6),
  count = c(745, 19987071, 2666, 20595360, 37737, 172006362, 80488, 174504898, 212258, 1272915272, 213766, 1280428583)
)

table3 <- tibble(
  country = rep(c("Afghanistan", "Brazil", "China"), each = 2),
  year = rep(c(1999, 2000), times = 3),
  rate = c("745/19987071", "2666/20595360", "37737/172006362", "80488/174504898", "212258/1272915272", "213766/1280428583")
)

table4a <- tibble(
  country = c("Afghanistan", "Brazil", "China"),
  `1999` = c(745, 37737, 212258),
  `2000` = c(2666, 80488, 213766)
)

table4b <- tibble(
  country = c("Afghanistan", "Brazil", "China"),
  `1999` = c(19987071, 172006362, 1272915272),
  `2000` = c(20595360, 174504898, 1280428583)
)
```

### 12.2.1 

**1. Using prose, describe how the variables and observations are organised in each of the sample tables.**  
*only need to think about this question*  
`table1`: tidy  
`table2`: one observation per row, but more than one variable per column  
`table3`: more than one value per cell  
`table4a` and `table4b`: more than one observation per row  

**2. Compute the `rate` for `table2`, and `table4a` + `table4b`. You will need to perform four operations:**

  1. Extract the number of TB cases per country per year.
  2. Extract the matching population per country per year.
  3. Divide cases by population, and multiply by 10000.
  4. Store back in the appropriate place.
  
  **Which representation is easiest to work with? Which is hardest? Why?**
  
```{r}
# rate for table2

# extract cases per country per year
(cases <- table2 %>% filter(type == "cases"))
# extract population per country per year
(population <- table2 %>% filter(type == "population"))

# divide cases by population and multiply by 10000
(rate <- tibble(
  country = rep(c("Afghanistan", "Brazil", "China"), each = 2),
  year = rep(c(1999, 2000), times = 3),
  type = "rate", 
  count = cases$count / population$count * 10000
))

# store back in table2
table2_rate <- rbind(table2, rate) %>% 
  arrange(country, year, type)
table2_rate
```

```{r}
# rate for table4a + table4b

# extract cases per country per year
## this is what table4a is
# extract population per country per year
## this is what table4b is

# divide cases by population and multiply by 10000
(rate <- tibble(
  country = rep(c("Afghanistan", "Brazil", "China"), each = 2),
  year = rep(c(1999, 2000), times = 3),
  type = "rate", 
  count = cases$count / population$count * 10000
))

# store back in appropriate place
## each variable is stored in its own tibble, so this fits the current format
```

I suppose the `table4` representation is easier to work with, while the `table2` representation is harder to work with. This is because `table4` already had the variables split up, so I didn't have to separate out cases and population like I had to do with `table2`. However, `table1` would have certainly been the easiest.

**3. Recreate the plot showing change in cases over time using table2 instead of table1. What do you need to do first?**

```{r}
table2 %>% filter(type == "cases") %>% 
  ggplot(aes(year, count)) + 
  geom_line(aes(group = country), color = "grey50") + 
  geom_point(aes(color = country))
```

When using `table2`, you first need to filter the data to only include `cases` and not `population`. 

## 12.3 Spreading and gathering

### 12.3.3  

**1. Why are `gather()` and `spread()` not perfectly symmetrical? Carefully consider the following example:**  
```{r}
stocks <- tibble(
  year   = c(2015, 2015, 2016, 2016),
  half  = c(   1,    2,     1,    2),
  return = c(1.88, 0.59, 0.92, 0.17)
)
stocks %>% 
  spread(year, return) %>%
  gather("year", "return", `2015`:`2016`)
```

*(Hint: look at the variable types and think about column names.)*

**Both `spread()` and `gather()` have a convert argument. What does it do?**

When `stocks` is `spread()` and the years become column names, information about `year` variable type is lost. As a result, `year` is a character vector after `spread()` and `gather()`. 

`convert` appears to re-assign variable types according to R's best guess based on the values. This seems to produce more consistent variable types before and after `spread()` or `gather()`. 

**2. Why does this code fail?**

```{}
table4a %>% 
  gather(1999, 2000, key = "year", value = "cases")
#> Error in inds_combine(.vars, ind_list): Position must be between 0 and n
```

Since this is missing the backticks around each year, R is looking for column position 1999 and 2000. However, it can't find these positions because the dataset is not that large. 

**3. Why does spreading this tibble fail? How could you add a new column to fix the problem?**

```{}
people <- tribble(
  ~name,             ~key,    ~value,
  #-----------------|--------|------
  "Phillip Woods",   "age",       45,
  "Phillip Woods",   "height",   186,
  "Phillip Woods",   "age",       50,
  "Jessica Cordero", "age",       37,
  "Jessica Cordero", "height",   156
)
```

Spreading this tibble fails because not all of the name by key combinations are unique. When R tries to spread, it finds two rows for "Phillip Woods" and "age". A new column could be added with an assigned individual number or some other unique identification per person. This would distinguish between the two entries for "Phillip Woods". 

**4. Tidy the simple tibble below. Do you need to spread or gather it? What are the variables?**

```{r}
preg <- tribble(
  ~pregnant, ~male, ~female,
  "yes",     NA,    10,
  "no",      20,    12
)
```

This tibble needs to be gathered in order to be tidy.  
The variables are:
  * `pregnant`: "yes" or "no"
  * `sex`: "male" or "female"
  * `count`: number of pregnant people per sex

```{r}
preg %>% gather("male", "female", key = sex, value = count)
```

## 12.4 Separating and uniting

### 12.4.3

**1. What do the extra and fill arguments do in separate()? Experiment with the various options for the following two toy datasets.**  

The `extra` argument determines what happens when there are more separations that need to happen than the number specified in the `sep` vector. By default, R will drop the extra values and give you a warning. However, you can tell R to either `drop` extra values without a warning or `merge` extra values together while keeping them in the final dataset. 

```{r}
# default, drop with warning
tibble(x = c("a,b,c", "d,e,f,g", "h,i,j")) %>% 
  separate(x, c("one", "two", "three"), extra = "merge")

# merge
tibble(x = c("a,b,c", "d,e,f,g", "h,i,j")) %>% 
  separate(x, c("one", "two", "three"), extra = "merge")
```

The `fill` argument conversely determines what happens when the `sep` vector is longer than the number of separations that need to happen. Similar to `extra`, the default option is to fill from the right and give you a warning. However, you can tell R to omit the warning and either fill values from the `right` or `left`. 

```{r}
# default, fill from right with warning
tibble(x = c("a,b,c", "d,e", "f,g,i")) %>% 
  separate(x, c("one", "two", "three"))

# fill from left
tibble(x = c("a,b,c", "d,e", "f,g,i")) %>% 
  separate(x, c("one", "two", "three"), fill = "left")
```

**2. Both `unite()` and `separate()` have a `remove` argument. What does it do? Why would you set it to `FALSE`?**  

By default, `unite()` and `separate()` don't include the initial converted column(s) in the output, which is the same as `remove = TRUE`. So generally, `remove` determines if the input column(s) are retained in the output or not. If you would like to keep the input column in the output, then you would set `remove = FALSE`. 

```{r}
tibble(x = c("a,b,c", "d,e,f", "g,h,i")) %>% 
  separate(x, c("one", "two", "three"))

tibble(x = c("a,b,c", "d,e,f", "g,h,i")) %>% 
  separate(x, c("one", "two", "three"), remove = FALSE)
```

Personally, I've done this before when I wanted to have a column containing two variables to `color` by, but then I wanted to `facet` by only one of those variables. For instance separate "light_treatment" by `color` and facet by "light". 

**3. Compare and contrast `separate()` and `extract()`. Why are there three variations of separation (by position, by separator, and with groups), but only one unite?**   

Both `separate()` and `extract()` can split one column into multiple output columns. However, `separate()` can only split by position or a separator while `extract()` can use regular expressions to determine groups and split by those groups. 

There are three variations of separation because there are a lot of possibilies for how you would want to split one column, but there is only one unite because there is really only one output method for joining columns together.

## 12.6 Case Study

### 12.6.1 

```{r}
# recreate tidy data
who1 <- who %>%
  gather(key, value, new_sp_m014:newrel_f65, na.rm = TRUE) %>% 
  mutate(key = stringr::str_replace(key, "newrel", "new_rel")) %>%
  separate(key, c("new", "var", "sexage")) %>% 
  select(-new) %>% 
  separate(sexage, c("sex", "age"), sep = 1)
```

**3. I claimed that iso2 and iso3 were redundant with country. Confirm this claim.**

```{r}
head(who1)

# pull out unique rows and count by country
who1[ , 1:3] %>%
  unique() %>%
  count(country) %>% 
  filter(n > 1)
```

Each country only has one row in this unique dataset, which means that every country has the same iso2 and iso3 entries. This makes sense because the data documentation specifies that iso2 contains two letter ISO country codes and iso3 containes three letter ISO country codes, so they should be redundant. 

**4. For each country, year, and sex compute the total number of cases of TB. Make an informative visualisation of the data.**

```{r}
# remove iso2 and iso3 columns
who2 <- who1 %>% 
  select(-iso2, -iso3)

# convert into factors
who2$country <- as.factor(who2$country)
who2$sex <- as.factor(who2$sex)
  
# compute total cases of TB for each country, year, and sex
who3 <- who2 %>% 
  group_by(country, year, sex) %>% 
  summarize(cases = sum(value))
head(who3)
  
# visualize 
who3 %>% 
  ggplot(aes(x = year, y = cases, color = country)) + 
  facet_wrap(~ sex) + 
  theme(legend.position = "none") + 
  geom_point() 

who3 %>% 
  ggplot(aes(year, cases, color = country)) + 
  facet_wrap(~ sex) + 
  theme(legend.position = "none") + 
  geom_line() + 
  geom_point()

# try zooming in
who3 %>% 
  ggplot(aes(year, cases, color = country)) + 
  facet_wrap(~ sex) + 
  theme(legend.position = "none") + 
  geom_line() + 
  geom_point() + 
  coord_cartesian(ylim = c(0, 2e5))
```

