---
title: "Swirl Notes"
author: "Cassandra"
date: "July 10, 2019"
output: 
  html_document: 
    keep_md: yes
---

```{r setup, include=FALSE}
library(swirl)
library(tidyverse)
# name = calbaker
```

Use `bye()` to exit swirl and save progress.   
Use `info()` to display options for swirl.

## R Programming

### 1. Basic Building Blocks

Values can be assigned into variables and stored for later use. 
```{r}
x <- 5 + 7
x
```

Vectors of numbers can also be stored in R.
```{r}
z <- c(1.1, 9, 3.14)
z
```

Vectors can then be combined into new vectors or used for calculations.
```{r}
c(z, 555, z)
z*2 + 100
```

Common operators  

* `+` = addition  
* `-` = subtraction  
* `*` = multiplication  
* `/` = division  
* `^` = exponents  
* `sqrt()` = square root  
* `abs()` = absolute value   

**Note:** When working with vectors, R will perform operations element by element. If two vectors are different lengths though, R will recycle the shorter vector. 
```{r}
long <- c(1, 2, 3, 4)
short <- c(0, 10)

long + short
```

### 3. Sequences of Numbers

Creating a basic sequence of numbers: 
```{r}
1:20 
seq(1, 20)
```

`seq` can also specify increments other than the default of 1.
```{r}
seq(0, 10, by = 0.5)
```

Make a sequence of a certain length:
```{r}
my_seq <- seq(5, 10, length = 30)
length(my_seq)
```

Make a sequence the length of another vector:
```{r}
1:length(my_seq)
seq(along.with = my_seq)
seq_along(my_seq)
```

create vectors using `rep()`: 
```{r}
# repeat something in order a specific number of times
rep(0, times = 40)
rep(c(0, 1, 2), times = 10)

# repeat each thing a specific number of times
rep(c(0, 1, 2), each = 10)
```

### 4. Vectors

Atomic vectors contain one data type while lists may contain multiple data types.
```{r}
num_vect <- c(0.5, 55, -10, 6)
(tf <- num_vect < 1)
```

Logical operators

* `<` = less than
* `>` = greater than
* `<=` = less than or equal to
* `>=` = greater than or equal to
* `==` = equal to 
* `!=` = not equal to 
* `|` = or
* `&` = and
* `!` = not 

Use `"` to notate characters.
```{r}
my_char <- c("My", "name", "is")
```

Vectors can also be collapsed or manipulated later.
```{r}
# combine multiple vector elements into one element
paste(my_char, collapse = " ")

# add a new element to the original vector
my_name <- c(my_char, "calbaker")
# and combine again 
paste(my_name, collapse = " ")
```

Use `paste` to join multiple vectors together. Recycling will occur here as well.
```{r}
paste(1:3, c("X", "Y", "Z"), sep = "")
paste(LETTERS, 1:4, sep = "-")
```

### 5. Missing Values 

`NA` represents missing or unavailable values. Operations involving `NA` will typically return `NA`. 
```{r}
x <- c(44, NA, 5, NA)
x*3
```

Additional practice with `NA`:
```{r}
# draw 1000 numbers from normal distribution
y <- rnorm(1000)
# make vector containing 1000 NAs
z <- rep(NA, 1000)
# randomly select 100 elements from the two vectors above
my_data <- sample(c(y, z), 100)

# where are the NA values?
(my_na <- is.na(my_data))
# how many NA values?
sum(my_na)
```

`NaN` is another notation in R that stands for 'not a number'. 
```{r}
0/0
```

### 6. Subsetting Vectors

Useful for only looking at specific elements of a vector. 
```{r}
x <- sample(c(y, z), 20)

# look at first ten elements of x
x[1:10]

# only select elements that aren't NA
y <- x[!is.na(x)]
y
```

Subsetting can be performed to fit multiple criteria at once. 
```{r}
# subset for values that are positive and not NA
x[!is.na(x) & x > 0]

# pull out specific elements from a vector
x[c(3, 5, 7)]

# exclude specific elements from a vector
x[c(-2, -10)]
x[-c(2, 10)]
```

Elements within vectors can be named.
```{r}
# create vector with names
vect <- c(foo = 11, bar = 2, norf = NA)
vect
# see the names of a vector
names(vect)

# or you can add names later
# first create vector without names
vect2 <- c(11, 2, NA)
# then add names
names(vect2) <- c("foo", "bar", "norf")

# same vector? 
identical(vect, vect2)

# subset using names
vect[c("foo", "bar")]
```

### 7. Matrices and Data Frames

Matrices can only have one data class while data frames can have many different data classes. 

Matrices can be created from vectors.
```{r}
# create initial vector
my_vector <- 1:20
dim(my_vector)

# change dimensions of vector 
dim(my_vector) <- c(4, 5)
# check new dimensions
dim(my_vector)
attributes(my_vector)

# this is now a matrix
class(my_vector)
my_matrix <- my_vector
```

`matrix()` can also be used to create matrices.
```{r}
my_matrix2 <- matrix(1:20, 4, 5)
identical(my_matrix, my_matrix2)
```

Additional information can be added to matrices later.
```{r}
# create name vector
patients <- c("Bill", "Gina", "Kelly", "Sean")

# combine name vector with information in matrix
cbind(patients, my_matrix)
```
**Note:** Since matrices can only have one class of data, R coerced the numbers into characters when the character vector was combined with the matrix.

Data frames must be used when there are multiple data classes involved.
```{r}
# combine name vector and matrix information into data frame
# make the data frame
my_data <- data.frame(patients, my_matrix)
my_data
# check that this is a data frame
class(my_data)
```

Additional information can also be added to data frames later.
```{r}
# create a vector of column names
cnames <- c("patient", "age", "weight", "bp", "rating", "test")

# assign the column names to the data frame
colnames(my_data) <- cnames
my_data
```

### 8. Logic 

Logical expressions evaluate to `TRUE` or `FALSE`.   
Remember the logical operators from section 4 above. 
```{r}
TRUE == TRUE 
6 < 7
10 <= 10
5 != 7
```

These expressions can also be nested together to get an overall answer.
```{r}
(FALSE == TRUE) == FALSE
```

Using `!` means 'not' and will reverse the answer. 
```{r}
!(5 == 7)
```

For 'and' statements, use `&` to evaluate across an entire vector and `&&` to only evaluate the first element of a vector.
```{r}
TRUE & c(TRUE, FALSE, FALSE) 
TRUE && c(TRUE, FALSE, FALSE)
```

Similarly, `|` evaluates 'or' across an entire vector and `||` evalutes 'or' for only the first element of a vector.
```{r}
FALSE | c(TRUE, FALSE, FALSE)
FALSE || c(TRUE, FALSE, FALSE)
```

### 12. Looking at Data

R has many functions to help you get a sense of your data.
```{r}
# look at dimensions (rows then columns)
dim(mpg)

# number of rows
nrow(mpg)
# number of columns
ncol(mpg)

# look at variable names
names(mpg)

# preview the first few rows of data (6 by default)
head(mpg)
# look at the last ten rows of data
tail(mpg, 10)
```

`summary` is good for getting an overview of the data.
```{r}
summary(mpg)

# to look at categorical variables that were cut off 
table(mpg$class)
```

`str` also provides a concise view of the data structure.
```{r}
str(mpg)
```

