---
title: "07.12 Meeting Notes"
author: "Cassandra"
date: "July 12, 2019"
output: 
  html_document: 
    keep_md: yes
---

```{r setup, message = FALSE}
library(tidyverse)
```

##Test header
no space between `#` and header text:   
Renders correctly when you knit in R, but doesn't render correctly online in Github.

## Header  
space between `#` and header text:   
Renders correctly when you knit in R and also online in Github.

**Other formatting notes**   
If you don't skip a line after code chunks, your text will start on the same line as your graph in the final markdown file.  
Also remember that if you want something to start on a different line than the text before it, you need to introduce a double space `  ` at the end of the preceding line. 

**Shortcut Tips:**   
`Ctrl` + `Alt` + `i` to make new code chunk.   
`Alt` + `-` to quickly make assignment operator `<-`.

**Uploading figure files to Github**
```{r}
ggplot(mpg) + 
  geom_point(aes(displ, hwy))
```

If a figure is included in a file, you need to also include the figures when you commit and push to Github. After you knit the file, you should see an additional item pop up in your Git tab in R Studio. It will probably be named something like `"your_file_name"_files`. When you click that box to commit and push, R will include all figures for that file. 

**Pulling in Github**   
This is only required when working with someone else on a file together. "Pulling" brings everything down from Github onto your computer to make sure you have the most recent version of the file on your computer, including any new changes from your collaborator. 

**Parts of a code chunk**  
If you exlude the `{r}` from the basic code chunk setup, it will format like a code chunk but R won't run the code when you knit the file. 
```
# will look like code chunk
# R won't run this though
ggplot(mpg) + 
  geom_point(aes(displ, hwy))
```

Use `#` to add comments to your code. This is useful for adding notes about your code and also for making R ignore troublesome lines of code. 
```{r}
# trying to make specific graph 
ggplot(mpg) + 
  geom_point(aes(displ, hwy)) #+ 
  #geom_smooth()
```

**Graphing specifications**   
Explicit calls like `data =` or `mapping =` aren't required as long as this information is in the order R is expecting. You can check what R expects by pressing `Tab` when your cursor is within a function or by looking at the help file.   
If you want to include information in a different order from what R expects, then you need to specify what each component is. 
```{r}
ggplot(mapping = aes(x = displ, y = hwy), data = mpg) + 
  geom_point()
```

